---
title: "Project0: Data exploration"
author: "Mahfuza Haque Mahi"
date: "2026-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(lme4)
library(ggplot2)
raw_dat <- read.csv(here::here("Project0/DataRaw", "Project0_Clean_v2.csv"))
```

## Initial overview of the dataset

```{r}
# Check structure
str(raw_dat)
head(raw_dat)
nrow(raw_dat) 

summary(raw_dat)
glimpse(raw_dat)
```

## View the Missingness pattern of variables:

```{r}
naniar::gg_miss_var(raw_dat)
naniar::vis_miss(raw_dat)
```

The investigator said the sample.interval variables may be incorrect and may not be used for inference. So, they will be removed them from analysis.


## Data Cleaning and construction of the analytic dataset

```{r}
# -----------------------------
# Construct Analytic Dataset
# -----------------------------

analysis_dat <- raw_dat %>%
  
  # Remove sample interval variables (deemed unreliable by investigator)
  # and hormone measures in alternative units to avoid redundancy
  select(
    -Booklet..Sample.interval,
    -Booklet..Sample.interval.Decimal.Time..mins.,
    -MEMs..Sample.interval,
    -MEMs..Sample.interval.Decimal.Time..mins.,
    -Cortisol..ug.dl.,
    -DHEA..pg.dl.
  ) %>%
  
  # Identify participants with repeated DHEA values at the assay
  # upper detection limit (5.205 nmol/L)
  group_by(SubjectID) %>%
  mutate(
    dhea_upper_hits = sum(DHEA..nmol.L. == 5.205, na.rm = TRUE)
  ) %>%
  ungroup()

# Identify participants to exclude:
# participants with more than two DHEA values at the detection limit
removed_info <- analysis_dat %>%
  filter(dhea_upper_hits > 2) %>%
  distinct(SubjectID, dhea_upper_hits)

participants_to_remove <- removed_info$SubjectID

# Apply participant-level exclusions and hormone quality-control rules
analysis_dat <- analysis_dat %>%
  
  # Exclude flagged participants entirely
  filter(!SubjectID %in% participants_to_remove) %>%
  
  # Apply biologically motivated hormone exclusions:
  #   - Cortisol > 80 nmol/L treated as implausible (set to missing)
  #   - DHEA at detection limit removed to avoid flattening diurnal patterns
  mutate(
    Cortisol..nmol.L. = ifelse(Cortisol..nmol.L. > 80, NA, Cortisol..nmol.L.),
    DHEA..nmol.L.     = ifelse(DHEA..nmol.L. == 5.205, NA, DHEA..nmol.L.)
  ) %>%
  
  # Remove temporary detection-limit counter
  select(-dhea_upper_hits) %>%
  
  # Clean clock-time variables:
  # trim whitespace and convert empty strings to missing values
  mutate(
    Booket..Clock.Time = na_if(trimws(Booket..Clock.Time), ""),
    MEMs..Clock.Time   = na_if(trimws(MEMs..Clock.Time), ""),
    Sleep.Diary.reported.wake.time = na_if(trimws(Sleep.Diary.reported.wake.time), "")
  ) %>%
  
  # Convert clock-time variables into POSIXct datetimes by
  # combining with the collection date
  mutate(
    booklet_time = as.POSIXct(
      paste(Collection.Date, Booket..Clock.Time),
      format = "%m/%d/%Y %H:%M",
      tz = "America/Denver"
    ),
    cap_time = as.POSIXct(
      paste(Collection.Date, MEMs..Clock.Time),
      format = "%m/%d/%Y %H:%M",
      tz = "America/Denver"
    ),
    reported_wake_time = as.POSIXct(
      paste(Collection.Date, Sleep.Diary.reported.wake.time),
      format = "%m/%d/%Y %H:%M",
      tz = "America/Denver"
    )
  ) %>%
  
  # Order observations to ensure correct propagation of wake time
  arrange(SubjectID, Collection.Date, Collection.Sample) %>%
  
  # Carry forward reported wake time within each subject and collection day
  group_by(SubjectID, Collection.Date) %>%
  fill(reported_wake_time, .direction = "down") %>%
  ungroup() %>%
  
  # Compute derived timing variables (in minutes):
  #   - Time since waking using booklet time
  #   - Time since waking using MEMS cap time
  #   - Difference between booklet and MEMS times (agreement metric)
  mutate(
    wake_diff_booklet = as.numeric(
      difftime(booklet_time, reported_wake_time, units = "mins")
    ),
    wake_diff_cap = as.numeric(
      difftime(cap_time, reported_wake_time, units = "mins")
    ),
    diff_booklet_cap = as.numeric(
      difftime(booklet_time, cap_time, units = "mins")
    )
  )
```

Data were cleaned and processed prior to analysis to remove timing variables deemed unreliable, exclude biologically implausible cortisol values (>80 nmol/L), and handle DHEA measurements at the assay detection limit. Participants with repeated DHEA values at the detection limit were excluded to avoid distortion of diurnal hormone patterns. These preprocessing steps ensured that the analytic dataset reflected valid measurements and aligned with the investigator’s prespecified quality-control criteria.

## Table 1

Table 1 is constructed using the analytic dataset after application of all prespecified quality-control and exclusion criteria, ensuring consistency between descriptive summaries and inferential analyses.

```{r}
# --- Minimal cleaning: make group + indicators readable ---
df_tab <- analysis_dat %>%
  mutate(
    # Grouping variable
    Collection.Sample = factor(
      Collection.Sample,
      levels = 1:4,
      labels = c("Waking", "+30 min", "Lunch", "+10 hours")
    ),


    # Time availability indicators (for missingness reporting)
    booklet_time_avail = ifelse(is.na(Booket..Clock.Time), "No", "Yes"),
    mems_time_avail    = ifelse(is.na(MEMs..Clock.Time), "No", "Yes"),
    wake_time_avail    = ifelse(is.na(Sleep.Diary.reported.wake.time), "No", "Yes"),

  ) %>%
  mutate(
    booklet_time_avail = factor(booklet_time_avail, levels = c("Yes","No")),
    mems_time_avail    = factor(mems_time_avail,    levels = c("Yes","No")),
    wake_time_avail    = factor(wake_time_avail,    levels = c("Yes","No"))
  )

# --- nice labels in the printed table ---
label(df_tab$DAYNUMB) <- "Study day (1–3)"
label(df_tab$booklet_time_avail) <- "Booklet time recorded"
label(df_tab$mems_time_avail) <- "MEMS cap time recorded"
label(df_tab$wake_time_avail) <- "Sleep diary wake time recorded"
label(df_tab$Cortisol..nmol.L.) <- "Cortisol (nmol/L)"
label(df_tab$DHEA..nmol.L.) <- "DHEA (nmol/L)"


# --- Table 1: overall + by sample type ---
table1(
  ~ booklet_time_avail + mems_time_avail + wake_time_avail +
    Cortisol..nmol.L. + DHEA..nmol.L.
  | Collection.Sample,
  data = df_tab,
  caption = "Table 1. Hormone measurements and timing data availability by collection time point."
)
```

Table 1 is stratified by collection sample (waking, +30 minutes, lunch, and +10 hours) because timing accuracy, missingness, protocol adherence, and hormone concentrations are expected to vary systematically across the day. Stratifying by sample preserves the biological interpretation of cortisol and DHEA measurements and allows clear assessment of protocol performance at each scheduled collection time.

### Looking at the outcome variables for question 3

```{r}
# Raw cortisol
ggplot(analysis_dat, aes(x = Cortisol..nmol.L.)) +
  geom_histogram(bins = 30, fill = "grey70", color = "black") +
  labs(
    title = "Histogram of raw cortisol values",
    x = "Cortisol (nmol/L)",
    y = "Count"
  )

# Log cortisol
ggplot(analysis_dat, aes(x = log(Cortisol..nmol.L.))) +
  geom_histogram(bins = 30, fill = "grey70", color = "black") +
  labs(
    title = "Histogram of log-transformed cortisol values",
    x = "log(Cortisol)",
    y = "Count"
  )

# Raw DHEA
ggplot(analysis_dat, aes(x = DHEA..nmol.L.)) +
  geom_histogram(bins = 30, fill = "grey70", color = "black") +
  labs(
    title = "Histogram of raw DHEA values",
    x = "DHEA (nmol/L)",
    y = "Count"
  )

# Log DHEA
ggplot(analysis_dat, aes(x = log(DHEA..nmol.L.))) +
  geom_histogram(bins = 30, fill = "grey70", color = "black") +
  labs(
    title = "Histogram of log-transformed DHEA values",
    x = "log(DHEA)",
    y = "Count"
  )

```
